\documentclass{llncs}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{url}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{float}

% Tighten spacing to fit 5 pages (no content removed)
\setlength{\parskip}{0.18em plus 0.06em minus 0.05em}
\addtolength{\textheight}{8em}
\addtolength{\topmargin}{-2.5em}
\addtolength{\headsep}{-0.5em}
\addtolength{\footskip}{-1em}

\title{Highly Dependable Systems 2025-2026
DepChain - Stage 1 - Group 39}
\author{Guilherme Ribeiro - 103562, Tomás Morais - 103619, João Ma - 102491}
\institute{Instituto Superior Técnico, University of Lisbon, Portugal}

\begin{document}
\maketitle
\vspace{-0.8em}
\begin{abstract}
This report presents the first stage of the DepChain project (SEC 2025--2026), focusing on the design and implementation of a simplified permissioned blockchain based on the Basic HotStuff BFT consensus algorithm. We describe the layered communication design (UDP, Fair Loss Link, Authenticated Perfect Links), the four-phase consensus protocol with explicit DECIDE round, the client library with signed requests to resist a Byzantine leader, and the mechanisms for crash and Byzantine fault tolerance. The report includes an explicit analysis of threats and corresponding protection mechanisms, and summarises the dependability guarantees provided by the system.
\end{abstract}
\vspace{-0.5em}
\section{Introduction}

DepChain is a permissioned (closed-membership) blockchain system with high dependability guarantees. Stage~1 focuses on the consensus layer using Basic HotStuff (Algorithm~2)~\cite{hotstuff}. The system comprises: (1) a \emph{client library} linked with the application, which translates \texttt{append(string)} calls into requests to the blockchain service; (2) \emph{blockchain members} (replicas) that run the BFT consensus and maintain an in-memory append-only log. Membership is static and known to all participants before system start; a Public Key Infrastructure (PKI) is assumed, with self-generated public/private keys pre-distributed in the membership. Communication is over UDP with no TLS; we build Authenticated Perfect Links on top of a best-effort (Fair Loss) layer, as required by the assignment.

\section{Design and Justification}
\vspace{-0.6em}
\subsection{Architecture and Components}

The implementation is structured in layered modules:
\begin{itemize}
  \setlength{\itemsep}{1pt}\setlength{\parsep}{0pt}\setlength{\topsep}{2pt}
  \item \textbf{Transport}: \texttt{UdpTransport} provides raw datagram send/receive; \texttt{FairLossLink} adds best-effort retries (limited number of retransmissions with delay) to approximate fair loss.
  \item \textbf{Links}: \texttt{AuthenticatedPerfectLink} (APL) provides authenticated, exactly-once delivery: each message is signed; on receive we verify the signature \emph{before} deduplication.
  \item \textbf{Config}: \texttt{Membership} holds static member IDs, network addresses, and public keys; $n$, $f = (n-1)/3$, and quorum size $2f+1$ are derived from the membership; leader for view $v$ is \texttt{memberIds.get(view mod n)}.
  \item \textbf{Consensus}: \texttt{HotStuffReplica} implements the four phases (PREPARE, PRE-COMMIT, COMMIT, DECIDE); \texttt{ConsensusNetwork} abstracts sending to all or to the leader (implemented by \texttt{APLConsensusNetwork} over APL).
  \item \textbf{Blockchain}: \texttt{BlockchainService} is an in-memory append-only list of strings; \texttt{BlockchainMember} wires the client listener, consensus replica, and the DECIDE upcall that applies blocks and sends responses to clients.
  \item \textbf{Client}: \texttt{DepChainClient} broadcasts signed append requests to all members and waits for a response; \texttt{ClientProtocol} defines the wire format and request/response parsing and verification.
\end{itemize}

\subsection{Layered Communication}

We use the following stack:
\begin{itemize}
  \setlength{\itemsep}{1pt}\setlength{\parsep}{0pt}\setlength{\topsep}{2pt}
  \item \textbf{UDP}: Raw datagram send/receive; messages may be lost, delayed, duplicated, or corrupted.
  \item \textbf{Fair Loss Link}: Sends the payload once and schedules a limited number of retries (e.g., 5 attempts with 40\,ms spacing). This approximates ``fair loss'': if the link is eventually reliable, the message will eventually be received. Duplicates may still occur; the layer above handles them.
  \item \textbf{Authenticated Perfect Links}: Each APL message carries \texttt{senderId}, \texttt{messageId}, and \texttt{payload}. The sender signs the tuple \texttt{(senderId, messageId, payload)} with RSA (SHA256withRSA). On receive we: (1) parse the wire; (2) verify that \texttt{senderId} is a member and obtain its public key; (3) \emph{verify the signature}; (4) \emph{then} deduplicate by \texttt{(senderId, messageId)}. We never deduplicate before authenticating: a Byzantine node could otherwise forge message IDs and cause legitimate messages to be discarded when they arrive~\cite{hotstuff}. A bounded ``seen'' set with pruning avoids unbounded memory growth.
\end{itemize}

\subsection{Basic HotStuff: Four Phases}

We implement the four phases of Algorithm~2. In PREPARE, the leader sends a proposal (block and highQC); replicas validate the block (see below), check the lock rule, and send a signed PREPARE\_VOTE; the leader collects $2f+1$ valid votes into a prepareQC. In PRE-COMMIT, the leader sends the block with prepareQC; replicas verify the QC (signatures and phase), update their lock, and send PRE-COMMIT\_VOTE; the leader builds preCommitQC. In COMMIT, the leader sends the block with preCommitQC; replicas verify and send COMMIT\_VOTE; the leader builds commitQC. The \emph{fourth phase (DECIDE)} is an explicit message round: the leader sends a DECIDE message containing the commitQC to all replicas; only when a replica receives DECIDE (or the leader, after sending it) do we invoke the DecideCallback (upcall), apply the block, and advance the view. This ensures a single, well-defined decision point. Votes are signed over \texttt{(view, phase, blockHash)}; the leader verifies each vote signature before counting it toward quorum, so up to $f$ Byzantine replicas cannot break safety.

\paragraph{Simplification with respect to the paper.} The HotStuff paper models proposals as nodes in a tree with a parent and a justify QC, and defines \texttt{safeNode(node, qc)} so that replicas only vote when the node extends the locked branch or the justify QC allows a safe unlock. We do \emph{not} implement that tree structure: our \texttt{Block} has no parent pointer and we do not maintain ancestry. The lock rule is implemented as a simplification: we keep a \texttt{lockedBlockHash} (set on pre-commit) and in PREPARE we only vote if the proposal matches the locked block or we accept the leader's justify QC when its view is below the current view (allowing unlock). This preserves safety in our setting (one proposal per view, no chaining as per the assignment) while keeping the implementation simpler; we state this explicitly so that the design is not overstated relative to the full Algorithm~2.

\subsection{Client--BFT Gap and Byzantine Leader Protection}

Clients expose \texttt{append(string)} and expect a confirmation (and index). The BFT layer exposes Propose and Decide. We fill the gap as follows. The client \emph{broadcasts} each append request to \emph{all} blockchain members (the client does not know the current leader). Each request includes \texttt{requestId}, the string, the client's public key, and a signature over \texttt{(requestId, string)}. Members receive requests on a dedicated client port (UDP); they verify the signature with the provided public key. Only \emph{verified} requests are stored in \texttt{verifiedClientRequests} and enqueued for the leader. The leader only proposes blocks whose payload is \texttt{requestId} plus string; replicas use a \texttt{BlockValidator} that checks that the block's payload corresponds to a previously verified client request. Thus a Byzantine leader cannot get correct replicas to decide on a forged command: they will reject the block in PREPARE. Invalid or malformed client messages (wrong type, oversized string, missing signature) are discarded by \texttt{ClientProtocol.parseRequest} and never trigger a proposal. On DECIDE, members apply the string to the log and send a response (requestId, success, index) to the client address stored when the request was first received.

\subsection{View Change and Pacemaker}

A timeout-based pacemaker triggers view change when the current view does not make progress. Each replica maintains \texttt{lastProgressTimeMs}; whenever it receives a relevant consensus message or sends a vote, it updates this timestamp. If \texttt{System.currentTimeMillis() - lastProgressTimeMs > viewTimeoutMs}, the replica increments the view, resets phase to PREPARE, clears votes and current block, and continues. The new leader is \texttt{getLeaderId(view)}; it uses \texttt{highQC} (the last commit QC) for the next PREPARE. Replicas that are behind in view catch up when they receive a message with a higher view. This handles crash faults of the leader.

\subsection{Byzantine Protection in Consensus}

We use per-replica digital signatures (RSA, Java Crypto API) for all votes; a Quorum Certificate is an aggregate of $2f+1$ such signatures, each verified with the corresponding member's public key from the static membership. Malformed or invalid votes are rejected; the leader only counts valid signatures toward quorum. With $n \ge 3f+1$, at most $f$ Byzantine replicas cannot form a quorum for a different value. Threshold signatures could be used for more compact QCs~\cite{weavechain} but are not required for correctness; we use aggregated individual signatures in this stage.

\section{Threat Analysis and Protection Mechanisms}
\vspace{-0.5em}
Table~\ref{tab:threats} summarises the main threats and the mechanisms used to address them.

\begin{table}[H]
  \caption{Threats and protection mechanisms.}
  \label{tab:threats}
  \centering
  \small
  \renewcommand{\arraystretch}{0.85}
  \begin{tabularx}{\linewidth}{@{}lXX@{}}
    \toprule
    \textbf{Threat} & \textbf{Description} & \textbf{Protection} \\
    \midrule
    Message forgery / impersonation & Adversary injects or alters messages & APL: verify signature with sender's public key; only verified messages delivered. Deduplication after verification. \\
    Byzantine replicas: invalid votes & Replicas send wrong or fake votes & Vote signatures verified; invalid votes ignored. Quorum requires $2f+1$ valid signatures. \\
    Byzantine leader: forged commands & Leader proposes block not from a client & BlockValidator: block payload must match a verified client request (requestId + string). Replicas reject in PREPARE. \\
    Leader crash & Leader stops proposing & Timeout-based pacemaker; view change; new leader continues with highQC. \\
    Network loss / delay & Messages dropped or delayed & Fair Loss Link retries; view change allows progress under partial synchrony. \\
    Replay / duplicate messages & Old messages replayed & APL deduplication by (senderId, messageId) after authentication. \\
    Invalid client requests & Malformed or oversized input & ClientProtocol: type, length, string size limit; parseRequest returns null; no proposal. \\
    \bottomrule
  \end{tabularx}
\end{table}

\subsection{Testing and Intrusive Mechanisms}
\vspace{-0.4em}
To test correctness beyond simple black-box load tests, we use \emph{intrusive} test harnesses that alter message delivery or replica behaviour:
\begin{itemize}
  \setlength{\itemsep}{1pt}\setlength{\parsep}{0pt}\setlength{\topsep}{2pt}
  \item \textbf{ByzantineTest}: One replica (e.g., replica 3) uses a \texttt{CorruptingConsensusNetwork} wrapper that corrupts outgoing vote messages (e.g., flips bits in the signature bytes). Correct replicas still decide: the leader collects only valid votes and reaches quorum with the other $2f+1$ replicas. This demonstrates detection and tolerance of Byzantine vote behaviour.
  \item \textbf{HotStuffIntegrationTest\#viewChangeAfterLeaderTimeout}: Replica 0 (leader) never proposes; after the view timeout, all replicas move to view 1; replica 1 becomes leader and proposes; all decide. This demonstrates crash handling via the pacemaker.
  \item \textbf{IntrusiveNetworkTest}: Uses \texttt{DropConsensusNetwork} to inject loss on a fraction of sends. With 0\% drop the test passes; the drop probability can be increased to observe resilience (retries and view change).
  \item \textbf{ClientIntegrationTest}: Client broadcasts append to all members; leader proposes; all decide and send responses; client receives confirmation.
\end{itemize}

These tests are run with JUnit (e.g. \texttt{mvn test -Dtest=ByzantineTest}). The README documents how to run the demos and tests.

\section{Dependability Guarantees}
\vspace{-0.5em}
\begin{itemize}
  \setlength{\itemsep}{1pt}\setlength{\parsep}{0pt}\setlength{\topsep}{2pt}
  \item \textbf{Safety}: Under the assumption that at most $f$ of $n \ge 3f+1$ replicas are Byzantine, all correct replicas decide the same sequence of blocks (agreement). The decided value is valid in the sense that it either comes from a verified client request (client blocks) or is an empty/no-op block (validity). The lock rule and QC verification ensure that no two different blocks are committed for the same view.
  \item \textbf{Liveness}: After Global Stabilization Time (GST), with a correct leader and the timeout-based pacemaker, the system eventually decides: either the current view progresses to DECIDE, or after timeout the view changes and the new leader (if correct) can make progress using highQC.
  \item \textbf{Authenticated communication}: APL provides integrity and origin authentication for all consensus messages. Client requests are verified (signature) before being stored and proposed; invalid messages are never proposed.
  \item \textbf{Resilience to Byzantine leader}: A Byzantine leader cannot force correct replicas to decide a forged client command, because replicas only vote for blocks that pass the BlockValidator (payload must match a verified client request).
\end{itemize}

\section{Conclusion}
\vspace{-0.5em}
DepChain Stage~1 implements Basic HotStuff with a strict four-phase protocol including an explicit DECIDE round, layered communication over UDP (Fair Loss Link then Authenticated Perfect Links with verify-then-deduplicate), client broadcast with signed requests and leader-based proposal handling, and protection against both crash and Byzantine faults. The design explicitly addresses the client--BFT gap and the threat of a Byzantine leader proposing forged commands. Intrusive tests demonstrate Byzantine vote detection and view change. The dependability guarantees (safety, liveness, authenticated communication) are aligned with the assignment requirements.

\begin{thebibliography}{9}
\bibitem{hotstuff} Yin, M., Malkhi, D., Reiter, M.K., Gueta, G.G., Abraham, I.: HotStuff: BFT Consensus in the Lens of Blockchain. arXiv:1803.05069 (2018).
\bibitem{weavechain} Weavechain threshold-sig: \url{https://github.com/weavechain/threshold-sig}.
\bibitem{lncs} Information for Authors of Springer Computer Science Proceedings (LNCS). \url{https://www.springer.com/gp/computer-science/lncs/conference-proceedings-guidelines}.
\end{thebibliography}
\end{document}
